---
title: "A stochastic meta-population model of LSD epidemics"
author: "Thibaut Jombart"
date: "`r Sys.Date()`"
output: github_document
---


This repository contains the __R__ package _lsdsim_ which implements a 
stochastic, meta-population compartmental model of Lumpy Skin Disease (LSD), 
including the simulation of different types of interventions.

It also contains an an instance of [_reportfactory_](https://www.reconverse.org/reportfactory/) to run epidemic simulations in a reproducible way. 

This repository is still under development. In particular, documentation and testing will be improved in early 2026. If you intend on using its content, do 
not hesitate to contact me at [thibautjombart@gmail.com](mailto:thibautjombart@gmail.com).


## Getting started

### Downloading the repository

You first need to clone or download the content of this repository, and open
an R session inside the main folder. 

To clone the repository, use:
```{bash, eval = FALSE}
git clone git@github.com:thibautjombart/lsdsim.git
```

If you have already cloned the repository, make sure you have the latest version
by opening a terminal in the main folder and typing:
```{bash, eval = FALSE}
git pull
```

If you use [Rstudio](https://posit.co/download/rstudio-desktop/), the easiest way to open a session is via opening the file _lsdv_simulations.Rproj_.

To install _lsdsim_ you will need to have _devtools_ installed on your computer
and to be able to compile R packages from the source. You will also need to have
the _reportfactory_ package installed. The latter will be handy to handle 
dependencies of the Rmarkdown documents stored in _report_sources/_. 


### Installing dependencies

The following code should install all needed packages:

```{r eval = FALSE}
## install and load lsdsim
if (!required(devtools)) install.packages("devtools")
devtools::load_all()

## install reportfactory, install Rmd deps
if (!required(reportfactory)) install.packages("reportfactory")
reportfactory::install_deps()

```


## Generating results

### Published results

Results presented in the initial publications are stored as:

- _outputs/res_2025-12-29.qs_: R object storing summaries of 6,000 simulations
- _outputs/simulations_final.html_: report of the 6,000 simulations


### Generating new results

The _reportfactory_ can be used to generate a report running simulations of epidemics with different types of responses, and storing the output in a time-stamped folder in _outputs/_. The sources of the report itself are in 
_report_sources/simulations.Rmd_. Users can modify this report as they see fit, 
or add a new report to the _report_sources_ folder to run new simulations.

The _simulations.Rmd_ report has the following parameters:

- n_rep: an `integer` indicating the number of replicates to generate; in our original study, we used 1,000, but beware that this will take hours to run on a standard desktop; defaults to 10
- run_sim: a `logical` indicating whether new simulations should be run; if `FALSE`, the latest simulation results stored inside `outputs` will be used to generate the report; defaults to `TRUE`

For instance, if you want to generate new results with 30 replicates, you can 
type:

```{r eval = FALSE}
reportfactory::compile_reports(params = list(n_rep = 30, run_sim = TRUE))
```
```{r eval = FALSE, results = "markup"}
>>> Compiling report: simulations
      - with parameters: n_rep = 30, run_sim = TRUE
All done!
```



## The _lsdsim_ package

The _lsdsim_ package was started around Christmas time to provide a platform for 
testing different responses to LSD epidemics, in the wake of mass culling 
imposed in France which led to protests and social unrest. In a nutshell: it 
works, all main features are unit-tested, but it is still under development. At this stage:

- unit-testing covers all core features including epidemic spread
and interventions; full test coverage will be ensured later
- documentation is largely lacking, but is underway
- CRAN checks are not passing (mostly due to the lack of documentation) but 
the package does install on current R versions
- a list of all core functions of the package will be provided here
- continuous integration will be implemented via github actions
- a dedicated website will be provided using _pkgdown_

