---
title: "A stochastic meta-population model of LSDV epidemics"
author: "Thibaut Jombart"
params:
  recompile: TRUE
output:
  html_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  collapse = TRUE, 
  fig.width = 8, 
  fig.height = 6, 
  dpi = 70, 
  cache = params$recompile
  )
```

## Outline of the work

In this report, we use a stochastic, meta-population SEIRDV model to simulate 
outbreaks of Lumpy skin disease virus. We contrast two types of response:

1. a baseline response which replicates current response in France with mass 
culling of infected herds

2. an alternative response which uses quarantine of infected herds, with no culling


## Simulation setup

### Required libraries

We load required libraries:
```{r}
library(tidyverse)
library(viridis)
library(qs)
#library(here)
library(compiler)
devtools::load_all()
```

## Population setting

Our benchmark suggests simulations for 365 days become prohibitively slow for
grid sizes exceeding 40 (_i.e._, 1600 farms). For the final simulation sets, we 
aim for 1000 replicates in each setting, so we will use a grid size of 40.
We assume a diffusion across neighbouring farms of 1%, _i.e._ 1% of transmission 
moves to a different farm nearby.

```{r}
grid_size <- 40
n_pop <- grid_size^2
xy <- make_grid(grid_size)
delta <- make_delta(xy, diffusion = 0.01)
```

The size of the farms, in number of herds, is drawn using an average herd size 
of 50, exponentially distributed with a minimum at 20 individuals:
```{r}
set.seed(1)
#herd_size <- round(rgamma(nrow(xy), shape = 18, scale = 50/18))
r_herd_size <- function(n) {
  20 + round(rexp(nrow(xy), 1/30))
}

herd_size <- r_herd_size(1e4)
hist(herd_size, border = "white", 
     main = "Herd size distribution",
     xlab = "Number of cows", 
     nclass = 30, 
     xlim = c(0, max(herd_size)))
summary(herd_size)
herds <- cbind.data.frame(xy, size = r_herd_size(n_pop))

ggplot(herds, aes(x = x, y = y)) + 
  geom_tile(aes(fill = size)) + 
  theme_bw() + 
  scale_fill_viridis_c("Herd size") + 
  labs(title = "Example of a meta-population")
```

In practice, every simulation will use a different population setting (randomly drawn herd sizes).

## Simulations

### General settings

We set the number of replicates, which will apply to each simulation scenario.
We also make a compiled version of `lsdsim` and set up parallelisation.

```{r}
n_rep <- 10
fast_lsdsim <- compiler::cmpfun(lsdsim)

```

This function is used to generate one new case in a randomly chosen population, 
to seed outbreaks:
```{r}
r_ini_case <- function(n_pop) {
  out <- rep(0, n_pop)
  out[sample.int(n_pop, 1)] <- 1
  out
}
```

Input parameters will be the same for different interventions:
```{r}
input_params <- rlsd_param(n_rep)
head(input_params)
```

Populations are drawn in advance as well, so the same populations will be used
in the different scenarios:
```{r}
list_herd_size <- lapply(1:n_rep, function(i) r_herd_size(n_pop))
```

Initial infections are also drawn in advance and will be common to all simulations:
```{r}
list_ini_cases <- lapply(1:n_rep, function(i) r_ini_case(n_pop))
```




## Baseline scenario: no response

In these simulations, we use no response.

```{r}

output_dir <- here::here("outputs", "simulations", "baseline")
dir.create(output_dir, recursive = TRUE)

for (i in seq_len(n_rep)) {
  ## run simulation
  current_sim <- fast_lsdsim(
    grid_size = grid_size, 
    ini_S = list_herd_size[[i]], 
    ini_I = list_ini_cases[[i]],
    delta = delta,
    beta = input_params[i, "beta"],
    sigma = input_params[i, "sigma"],
    gamma = input_params[i, "gamma"],
    quarantine = FALSE,
    vaccination = FALSE, 
    culling = FALSE, 
    insecticide = FALSE, 
    cfr = 0.03 # CFR reportedly 1-5%
  )
  
  ## save full simulation to file
  out_file <- file.path(output_dir, paste0("sim_", i, ".rds"))
  qsave(current_sim, out_file)
  
  ## store summary
  
}

```


## Impact of mass culling

Here we simulate outbreaks aiming to replicate the current setting:


## Visualising results

### Time series

Here we first sum compartments across all patches, and visualize the results as a time series:

```{r}
comps <- c("V", "E", "I", "F", "D", "C", "R")
res_smry <- list()
for(e in comps) {
  to_keep <- grep(sprintf("^[%s]", e), colnames(res))
  res_smry[[e]] <- rowSums(res[, to_keep])
}
res_smry <- cbind(time = 1:365, data.frame(res_smry))
head(res_smry)

res_smry %>%
  pivot_longer(-1, names_to = "Compartment", values_to = "n") %>% 
  ggplot(aes(x = time, y = n, color = Compartment)) + 
  geom_line(size = 1) + 
  theme_bw() +
  labs(
    x = "Time",
    y = "Number of individuals", 
    title = "Overall dynamics"
    ) + 
  scale_x_continuous(n.breaks = 10) + 
  scale_y_continuous(n.breaks = 10) +
  theme(legend.position = "bottom")
```


### Mapping total cases

Here we first isolate all compartments corresponding to infections, and calculate total cases for every patch at the end of the simulation. Then we map the results:

```{r fig.width = 5, fig.height = 5}
to_keep <- c(grep("^[EIDFRC]", colnames(res)))
temp <- res[res[, "step"] == 365, to_keep, drop = TRUE]
patch_id <- as.integer(gsub("[^[:digit:]]", "", names(temp)))
total_cases <- tapply(temp, patch_id, sum)

df <- cbind(patches, total_cases)

ggplot(df, aes(x = x, y = y)) + 
  geom_tile(aes(fill = total_cases)) + 
  scale_fill_viridis_c(
    option = "magma", 
    na.value = "transparent", 
    n.breaks = 7) + 
  labs(x = "Longitude", 
       y = "Latitude", 
       fill = "Total cases") + 
  scale_x_continuous(n.breaks = 10) + 
  scale_y_continuous(n.breaks = 10) + 
  theme(legend.position = "bottom")

```

