---
title: "A stochastic meta-population model of LSDV epidemics"
author: "Thibaut Jombart"
params:
  recompile: FALSE
output:
  html_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  collapse = TRUE, 
  fig.width = 8, 
  fig.height = 6, 
  dpi = 70
  )
```

## Outline of the work

In this report, we use a stochastic, meta-population SEIRDV model to simulate 
outbreaks of Lumpy skin disease virus. We contrast two types of response:

1. a baseline response which replicates current response in France with mass 
culling of infected herds

2. an alternative response which uses quarantine of infected herds, with no culling


## Simulation setup

### Required libraries

We load required libraries:
```{r}
library(tidyverse)
library(viridis)
library(qs)
library(here)
library(digest)
library(compiler)
devtools::load_all()
```

## Population setting

Our benchmark suggests simulations for 365 days become prohibitively slow for
grid sizes exceeding 40 (_i.e._, 1600 farms). For the final simulation sets, we 
aim for 1000 replicates in each setting, so we will use a grid size of 40.
We assume a diffusion across neighbouring farms of 1%, _i.e._ 1% of transmission 
moves to a different farm nearby.

```{r}
duration <- 365
grid_size <- 40
n_pop <- grid_size^2
xy <- make_grid(grid_size)
delta <- make_delta(xy, diffusion = 0.01)
```

The size of the farms, in number of herds, is drawn using an average herd size 
of 50, exponentially distributed with a minimum at 20 individuals:
```{r}
r_herd_size <- function(n) {
  20 + round(rexp(nrow(xy), 1/30))
}

herd_size <- r_herd_size(1e4)
hist(herd_size, border = "white", 
     main = "Herd size distribution",
     xlab = "Number of cows", 
     nclass = 30, 
     xlim = c(0, max(herd_size)))
summary(herd_size)
herds <- cbind.data.frame(xy, size = r_herd_size(n_pop))

ggplot(herds, aes(x = x, y = y)) + 
  geom_tile(aes(fill = size)) + 
  theme_bw() + 
  scale_fill_viridis_c("Herd size") + 
  labs(title = "Example of a meta-population")
```

In practice, every simulation will use a different population setting (randomly drawn herd sizes).

## Simulations

### General settings

We set the number of replicates, which will apply to each simulation scenario.
We also make a compiled version of `lsdsim` and set up parallelisation.

```{r}
n_rep <- 1e3
fast_lsdsim <- compiler::cmpfun(lsdsim)

```

This function is used to generate one new case in a randomly chosen population, 
to seed outbreaks:
```{r}
r_ini_case <- function(n_pop) {
  out <- rep(0, n_pop)
  out[sample.int(n_pop, 1)] <- 1
  out
}
```

Input parameters will be the same for different interventions:
```{r}
input_params <- rlsd_param(n_rep)
head(input_params)
```

Populations are drawn in advance as well, so the same populations will be used
in the different scenarios:
```{r}
list_herd_size <- lapply(1:n_rep, function(i) r_herd_size(n_pop))
```

Initial infections are also drawn in advance and will be common to all simulations:
```{r}
list_ini_cases <- lapply(1:n_rep, function(i) r_ini_case(n_pop))
```

We put all simulation inputs into a single `tibble`:
```{r}
inputs <- tibble(
  input_params, 
  ini_S = list_herd_size,
  ini_I = list_ini_cases
)

inputs$label <- apply(inputs, 1, function(e) substr(digest(e), 1, 8))

```



## Baseline scenario: no response

In these simulations, we use no response.

```{r baseline, cache = !params$recompile}
output_dir <- here::here("outputs", "simulations", "baseline", Sys.Date())
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}

res_baseline <- vector(mode = "list", length = n_rep)
for (i in seq_len(n_rep)) {
  ## run simulation
  current_sim <- fast_lsdsim(
    grid_size = grid_size, 
    time = duration,
    ini_S = unlist(inputs$ini_S[[i]]), 
    ini_I = unlist(inputs$ini_I[[i]]),
    delta = delta,
    beta = inputs$beta[i],
    sigma = inputs$sigma[i],
    gamma = inputs$gamma[i],
    cfr = inputs$cfr[i],
    quarantine = FALSE,
    vaccination = FALSE, 
    culling = FALSE, 
    insecticide = FALSE
  )
  
  ## save full simulation to file
  out_file <- file.path(
    output_dir, 
    paste0(inputs$label[i], "_baseline", ".qs")
  )
  qsave(current_sim, out_file)
  
  ## store summary
  res_baseline[[i]] <- summary(current_sim)[duration, ]
}

res_baseline <- cbind(label = inputs$label, bind_rows(res_baseline))
head(res_baseline)
summary(res_baseline)

qsave(res_baseline, file.path(output_dir, "summary.qs"))

```


