---
title: "A stochastic meta-population model of LSDV epidemics"
author: "Thibaut Jombart"
params:
  n_rep: 2
  run_sim: TRUE
output:
  html_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  collapse = TRUE, 
  fig.width = 8, 
  fig.height = 6, 
  dpi = 70
  )
```

## Outline of the work

In this report, we use a stochastic, meta-population SEIRDV model to simulate 
outbreaks of Lumpy skin disease virus. We contrast two types of response:

1. a baseline response which replicates current response in France with mass 
culling of infected herds

2. an alternative response which uses quarantine of infected herds, with no culling


## Simulation setup

### Required libraries

We load required libraries:
```{r}
library(tidyverse)
library(viridis)
library(qs)
library(here)
library(digest)
library(compiler)
library(rio)
library(cowplot)
devtools::load_all()
```

## Population setting

Our benchmark suggests simulations for 365 days become prohibitively slow for
grid sizes exceeding 40 (_i.e._, 1600 farms). For the final simulation sets, we 
aim for 1000 replicates in each setting, so we will use a grid size of 40.
We assume a diffusion across neighbouring farms of 1%, _i.e._ 1% of 
transmission moves to a different farm nearby.

```{r}
n_rep <- params$n_rep
duration <- 365
grid_size <- 40
n_pop <- grid_size^2
xy <- make_grid(grid_size)
delta <- make_delta(xy, diffusion = 0.05)
```

The size of the farms, in number of herds, is drawn using an average herd size 
of 50, exponentially distributed with a minimum at 20 individuals:
```{r}
r_herd_size <- function(n) {
  20 + round(rexp(nrow(xy), 1/30))
}

herd_size <- r_herd_size(1e4)
hist(herd_size, border = "white", 
     main = "Herd size distribution",
     xlab = "Number of cows", 
     nclass = 30, 
     xlim = c(0, max(herd_size)))
summary(herd_size)
herds <- cbind.data.frame(xy, size = r_herd_size(n_pop))

ggplot(herds, aes(x = x, y = y)) + 
  geom_tile(aes(fill = size)) + 
  theme_bw() + 
  scale_fill_viridis_c("Herd size") + 
  labs(title = "Example of a meta-population")
```

In practice, every simulation will use a different population setting (randomly drawn herd sizes).

## Simulations

### General settings

We set the number of replicates, which will apply to each simulation scenario.
We also make a compiled version of `lsdsim` and set up initial conditions.

```{r}
fast_lsdsim <- compiler::cmpfun(lsdsim)
```

This function is used to generate one new case in a randomly chosen population, 
to seed outbreaks:
```{r}
r_ini_case <- function(n_pop) {
  out <- rep(0, n_pop)
  out[sample.int(n_pop, 1)] <- 1
  out
}
```

Input parameters will be the same for different interventions:

```{r }
lsd_params <- rlsd_param(n_rep)
head(lsd_params)

list_herd_size <- lapply(1:n_rep, function(i) r_herd_size(n_pop))
list_ini_cases <- lapply(1:n_rep, function(i) r_ini_case(n_pop))

inputs <- tibble(
  lsd_params, 
  ini_S = list_herd_size,
  ini_I = list_ini_cases
)

inputs$label <- apply(inputs, 1, function(e) substr(digest(e), 1, 8))
```



### *baseline* scenario: no response

In these simulations, we use no response.

```{r baseline, eval = params$run_sim}
output_dir <- here::here("outputs", "simulations", "baseline", Sys.Date())
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}

res_baseline <- vector(mode = "list", length = n_rep)
for (i in seq_len(n_rep)) {
  ## run simulation
  current_sim <- fast_lsdsim(
    grid_size = grid_size, 
    time = duration,
    ini_S = unlist(inputs$ini_S[[i]]), 
    ini_I = unlist(inputs$ini_I[[i]]),
    delta = delta,
    beta = inputs$beta[i],
    sigma = inputs$sigma[i],
    gamma = inputs$gamma[i],
    cfr = inputs$cfr[i],
    quarantine = FALSE,
    vaccination = FALSE, 
    mass_culling = FALSE, 
    select_culling = FALSE,
    insecticide = FALSE
  )
  
  ## save full simulation to file
  out_file <- file.path(
    output_dir, 
    paste0(inputs$label[i], "_baseline", ".qs")
  )
  qsave(current_sim, out_file)
  
  ## store summary
  res_baseline[[i]] <- summary(current_sim)[duration, ]
}

res_baseline <- cbind(label = inputs$label, bind_rows(res_baseline))
head(res_baseline)
summary(res_baseline)

qsave(res_baseline, file.path(output_dir, "summary.qs"))
rio::export(res_baseline, file.path(output_dir, "summary.xlsx"))

```




### *cull_vacc* scenario: mass culling and ring vaccination

In these simulations, we use:

- mass culling of the affected herds
- ring vaccination (i.e. vaccination of neighbours)

```{r cull_vacc, eval = params$run_sim}
output_dir <- here::here("outputs", "simulations", "cull_vacc", Sys.Date())
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}

res_cull_vacc <- vector(mode = "list", length = n_rep)
for (i in seq_len(n_rep)) {
  ## run simulation
  current_sim <- fast_lsdsim(
    grid_size = grid_size, 
    time = duration,
    ini_S = unlist(inputs$ini_S[[i]]), 
    ini_I = unlist(inputs$ini_I[[i]]),
    delta = delta,
    beta = inputs$beta[i],
    sigma = inputs$sigma[i],
    gamma = inputs$gamma[i],
    cfr = inputs$cfr[i],
    quarantine = FALSE,
    vaccination = TRUE, 
    mass_culling = TRUE, 
    select_culling = FALSE,
    insecticide = FALSE,
    interv_delay = 7,
    interv_release = 28,
    vacc_coverage = 1, 
    vacc_efficacy = 0.65,
    rate_cull = 0.5
  )
  
  ## save full simulation to file
  out_file <- file.path(
    output_dir, 
    paste0(inputs$label[i], "_cull_vacc", ".qs")
  )
  qsave(current_sim, out_file)
  
  ## store summary
  res_cull_vacc[[i]] <- summary(current_sim)[duration, ]
}

res_cull_vacc <- cbind(label = inputs$label, bind_rows(res_cull_vacc))
head(res_cull_vacc)
summary(res_cull_vacc)

qsave(res_cull_vacc, file.path(output_dir, "summary.qs"))
rio::export(res_cull_vacc, file.path(output_dir, "summary.xlsx"))

```



### *quarant_vacc* scenario: quarantine and ring vaccination

In these simulations, we use:

- quarantine of infected herds and neighbours
- vaccination of infected herds and neighbours

```{r quarant_vacc, eval = params$run_sim}
output_dir <- here::here("outputs", "simulations", "quarant_vacc", Sys.Date())
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}

res_quarant_vacc <- vector(mode = "list", length = n_rep)
for (i in seq_len(n_rep)) {
  ## run simulation
  current_sim <- fast_lsdsim(
    grid_size = grid_size, 
    time = duration,
    ini_S = unlist(inputs$ini_S[[i]]), 
    ini_I = unlist(inputs$ini_I[[i]]),
    delta = delta,
    beta = inputs$beta[i],
    sigma = inputs$sigma[i],
    gamma = inputs$gamma[i],
    cfr = inputs$cfr[i],
    quarantine = TRUE,
    vaccination = TRUE, 
    mass_culling = FALSE, 
    select_culling = FALSE,
    insecticide = FALSE,
    interv_delay = 7,
    interv_release = 28,
    quarant_efficacy_in = 0.2,
    quarant_efficacy_out = 0.9,
    vacc_coverage = 1, 
    vacc_efficacy = 0.65
  )
  
  ## save full simulation to file
  out_file <- file.path(
    output_dir, 
    paste0(inputs$label[i], "_quarant_vacc", ".qs")
  )
  qsave(current_sim, out_file)
  
  ## store summary
  res_quarant_vacc[[i]] <- summary(current_sim)[duration, ]
}

res_quarant_vacc <- cbind(label = inputs$label, bind_rows(res_quarant_vacc))
head(res_quarant_vacc)
summary(res_quarant_vacc)

qsave(res_quarant_vacc, file.path(output_dir, "summary.qs"))
rio::export(res_quarant_vacc, file.path(output_dir, "summary.xlsx"))

```



### *selcull_quarant_vacc* scenario: selective culling, quarantine and ring vaccination

In these simulations, we use:

- selective culling of infected individuals
- quarantine of infected herds and neighbours
- vaccination of infected herds and neighbours
 

```{r selcull_quarant_vacc, eval = params$run_sim}
output_dir <- here::here("outputs", "simulations", "quarant_vacc", Sys.Date())
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}

res_selcull_quarant_vacc <- vector(mode = "list", length = n_rep)
for (i in seq_len(n_rep)) {
  ## run simulation
  current_sim <- fast_lsdsim(
    grid_size = grid_size, 
    time = duration,
    ini_S = unlist(inputs$ini_S[[i]]), 
    ini_I = unlist(inputs$ini_I[[i]]),
    delta = delta,
    beta = inputs$beta[i],
    sigma = inputs$sigma[i],
    gamma = inputs$gamma[i],
    cfr = inputs$cfr[i],
    quarantine = TRUE,
    vaccination = TRUE, 
    mass_culling = FALSE,
    select_culling = TRUE,
    insecticide = FALSE,
    interv_delay = 7,
    interv_release = 28,
    quarant_efficacy_in = 0.2,
    quarant_efficacy_out = 0.9,
    vacc_coverage = 1, 
    vacc_efficacy = 0.65, 
    rate_cull = 0.5
  )
  
  ## save full simulation to file
  out_file <- file.path(
    output_dir, 
    paste0(inputs$label[i], "_quarant_vacc", ".qs")
  )
  qsave(current_sim, out_file)
  
  ## store summary
  res_selcull_quarant_vacc [[i]] <- summary(current_sim)[duration, ]
}

res_selcull_quarant_vacc  <- cbind(
  label = inputs$label, 
  bind_rows(res_selcull_quarant_vacc )
)

head(res_selcull_quarant_vacc )
summary(res_selcull_quarant_vacc)

qsave(res_selcull_quarant_vacc, file.path(output_dir, "summary.qs"))
rio::export(res_selcull_quarant_vacc, file.path(output_dir, "summary.xlsx"))

```


### *all_but_cull* scenario: selective culling, quarantine, ring vaccination, insecticide (all but mass culling)

In these simulations, we use:

- selective culling of infected individuals
- quarantine of infected herds and neighbours
- vaccination of infected herds and neighbours
- insecticide spray of infected herds and neighbours


```{r all_but_cull, eval = params$run_sim}
output_dir <- here::here("outputs", "simulations", "all_but_cull", Sys.Date())
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}

res_all_but_cull <- vector(mode = "list", length = n_rep)
for (i in seq_len(n_rep)) {
  ## run simulation
  current_sim <- fast_lsdsim(
    grid_size = grid_size, 
    time = duration,
    ini_S = unlist(inputs$ini_S[[i]]), 
    ini_I = unlist(inputs$ini_I[[i]]),
    delta = delta,
    beta = inputs$beta[i],
    sigma = inputs$sigma[i],
    gamma = inputs$gamma[i],
    cfr = inputs$cfr[i],
    quarantine = TRUE,
    vaccination = TRUE, 
    mass_culling = FALSE,
    select_culling = TRUE,
    insecticide = TRUE,
    interv_delay = 7,
    interv_release = 28,
    vacc_coverage = 1, 
    vacc_efficacy = 0.65,
    quarant_efficacy_in = 0.2,
    quarant_efficacy_out = 0.9, 
    rate_cull = 0.5,
    insect_efficacy = 0.5
  )
  
  ## save full simulation to file
  out_file <- file.path(
    output_dir, 
    paste0(inputs$label[i], "_all_but_cull", ".qs")
  )
  qsave(current_sim, out_file)
  
  ## store summary
  res_all_but_cull[[i]] <- summary(current_sim)[duration, ]
}

res_all_but_cull <- cbind(label = inputs$label, bind_rows(res_all_but_cull))
head(res_all_but_cull)
summary(res_all_but_cull)

qsave(res_all_but_cull, file.path(output_dir, "summary.qs"))
rio::export(res_all_but_cull, file.path(output_dir, "summary.xlsx"))

```


### all scenario: mass culling, quarantine, ring vaccination, insecticide

In these simulations, we use mass culling, ring quarantine and vaccination, and insecticide spray.

```{r all, eval = params$run_sim}
output_dir <- here::here("outputs", "simulations", "all", Sys.Date())
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}

res_all <- vector(mode = "list", length = n_rep)
for (i in seq_len(n_rep)) {
  ## run simulation
  current_sim <- fast_lsdsim(
    grid_size = grid_size, 
    time = duration,
    ini_S = unlist(inputs$ini_S[[i]]), 
    ini_I = unlist(inputs$ini_I[[i]]),
    delta = delta,
    beta = inputs$beta[i],
    sigma = inputs$sigma[i],
    gamma = inputs$gamma[i],
    cfr = inputs$cfr[i],
    quarantine = TRUE,
    vaccination = TRUE, 
    mass_culling = TRUE, 
    select_culling = TRUE, # redundant with mass culling but for consistency
    insecticide = TRUE,
    interv_delay = 7,
    interv_release = 28,
    vacc_coverage = 1, 
    vacc_efficacy = 0.65,
    rate_cull = 0.5,
    quarant_efficacy_in = 0.2,
    quarant_efficacy_out = 0.9,
    insect_efficacy = 0.5
  )
  
  ## save full simulation to file
  out_file <- file.path(
    output_dir, 
    paste0(inputs$label[i], "_all", ".qs")
  )
  qsave(current_sim, out_file)
  
  ## store summary
  res_all[[i]] <- summary(current_sim)[duration, ]
}

res_all <- cbind(label = inputs$label, bind_rows(res_all))
head(res_all)
summary(res_all)

qsave(res_all, file.path(output_dir, "summary.qs"))
rio::export(res_all, file.path(output_dir, "summary.xlsx"))

```



## Analyses

```{r eval = params$run_sim}
res_baseline$scenario <- "baseline"
res_cull_vacc$scenario <- "cull_vacc"
res_quarant_vacc$scenario <- "quarant_vacc"
res_selcull_quarant_vacc$scenario <- "selcull_quarant_vacc"
res_all_but_cull$scenario <- "all_but_cull"
res_all$scenario <- "all"

res <- bind_rows(
  res_baseline, 
  res_cull_vacc, 
  res_quarant_vacc, 
  res_selcull_quarant_vacc,
  res_all_but_cull,
  res_all
  ) %>% 
  select(scenario, everything()) %>% 
  mutate(
    p_immun = (R + V) / N,
    p_dead = (D + C) / (N + D + C), 
    scenario = factor(
      scenario, levels = c(
        "baseline", 
        "cull_vacc", 
        "quarant_vacc",
        "selcull_quarant_vacc",
        "all_but_cull",
        "all")
      )
  )

file_name <- paste0("res_", Sys.Date(), ".qs")
qsave(res, here::here("outputs", file_name))

```

This is evaluated only if new results have not been obtained, fetching the 
latest version of the results:
```{r eval = !params$run_sim}
all_files <- list.files(here::here("outputs"), pattern = "^res")
extract_date <- function(x) {
  out <- gsub("[^[:digit:]-]", "", x)
  as.Date(out)
}
dates <- sapply(all_files, extract_date)
to_keep <- which.max(dates)
file_to_keep <- all_files[to_keep]
file_to_keep
message("reading results from file:", file_to_keep)
res <- qread(here::here("outputs", file_to_keep))
```



```{r out.width = "100%"}
label_numbers <- function(x, show_1_as_0 = FALSE) {
  out <- scales::label_number(scale_cut = scales::cut_long_scale())(x)
  if (show_1_as_0) {
    out[out=="1"] <- "0"
  }
  out
}

p_cases <- ggplot(res, aes(x = scenario, y = total_cases)) + 
  geom_violin(scale = "width") + 
  scale_y_continuous(
    transform = "log10", 
    labels = function(e) label_numbers(e, show_1_as_0 = TRUE)
  ) + 
  theme_bw() + 
  labs(
    y = "Total cases"
  )

p_n_pop <- ggplot(res, aes(x = scenario, y = n_pop_ever_infec)) + 
  geom_violin(scale = "width") + 
  scale_y_continuous(
    transform = "log10", 
    labels = function(e) label_numbers(e, show_1_as_0 = TRUE)
  ) + 
  theme_bw() + 
  labs(
    y = "Number of herds affected"
    )


p_immun <- ggplot(res, aes(x = scenario, y = p_immun)) + 
  geom_violin(scale = "width") + 
  theme_bw() + 
  labs(
    y = "Proportion of immunized\ncattles"
    )

p_dead <- ggplot(res, aes(x = scenario, y = (D+C))) + 
  geom_violin(scale = "width") + 
  scale_y_continuous(
    transform = "log10", 
    labels = function(e) label_numbers(e, show_1_as_0 = TRUE)
  ) + 
  theme_bw() + 
  labs(
    y = "Number of dead cattles"
    )

rotated_x <- theme(
  axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)
  )

plot_grid(
  p_cases + rotated_x, 
  p_n_pop + rotated_x,
  p_immun + rotated_x, 
  p_dead + rotated_x
)

```






